---
title: "Class08: Mini Project"
author: "Joli DeLucia (A17320476)"
format: pdf
toc: true
---

## Background and Preparing the Data

In today's class we will be employing all the R techniques for data analysis that we have learned thus far - including the machine learning methods of clustering and PCA - to analyze real breast cancer biopsy data. 

```{r}
wisc.df <- read.csv("WisconsinCancer.csv", row.names = 1)
```

We need to remove the `diagnosis` column before we do any further analysis of the dataset - we don't want to pass this to PCA, etc. We will save it as a separate vector that we can use later to compare ou findings to those of experts.

```{r}
head(wisc.df, 4)
wisc.data <- wisc.df[,-1]
wisc.data
```

# Create diagnosis vector for later 

```{r}
diagnosis <- wisc.df[,1]
```


> Q1. How many observations are in this dataset?

```{r}
nrow(wisc.df)
```

569 observations are in this dataset.

> Q2. How many of the observations have a malignant diagnosis?

```{r}
sum(wisc.df$diagnosis == "M")
```

There are 212 individuals that have a malignant diagnosis.

> Q3. How many variables/features in the data are suffixed with _mean?

```{r}
grep("_mean", colnames(wisc.df))
```

There are 10 columns that are suffixed with "_mean".

## Principal Component Analysis (PCA)

The main function in base R is called `prcomp()` we will use the optional arugement `scale = TRUE` here as the data columns/features/dimensions are on very different scales in the original data set.

```{r}
colMeans(wisc.data)
apply(wisc.data,2,sd)
```

```{r}
wisc.pr <- prcomp(wisc.data, scale = T)
summary(wisc.pr)
```

> Q4. From your results, what proportion of the original variance is captured by the first principal component (PC1)?

0.4427 is the proportion of the original variance captured in the first principal component.

> Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

3 PCs are required to describe at least 70% of the original variance in the data.

> Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?

7 PCs are required to describe at least 90% of the original variance in the data. 

## Interpreting PCA results

```{r}
biplot(wisc.pr)
```

> Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?

This plot is very difficultt to understand because there is a lot going on in such a little frame. Also the numbers and column names are preactically impossible to read due to the immense overlap and being in the same colors. It stands out to me that the numbers are very clustered but with some outlying numbers. 

```{r}
library(ggplot2)

ggplot(wisc.pr$x) +
  aes(PC1, PC2, col=diagnosis) +
  geom_point()
```

> Q8. Generate a similar plot for principal components 1 and 3. What do you notice about these plots?

```{r}
ggplot(wisc.pr$x) +
  aes(PC1, PC3, col = diagnosis) +
  geom_point()
```

Answer to Q8: I notice that the plot of PC1 vs PC3 is lower than in the plot of PC1 vs PC2. I notice that there is a separation between the blue, malignant patients and the red, benign patients.

# 4. Hierarchical Clustering

The goal of this section is to do hierarchical clustering of the original data to see if there is any obvious grouping into malignant and benign clusters. 

> Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean? This tells us how much this original feature contributes to the first PC. Are there any features with larger contributions than this one?

I did not see any features with larger contributions than this one with -0.26085376. 

```{r}
wisc.pr$rotation[,1]
```

```{r}
data.scaled <- scale((wisc.data))
data.dist <- dist(data.scaled)
wisc.hclust <- hclust(data.dist, method = "complete")
```

> Q10. Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters?

At height 19, there is 4 clusters. 

```{r}
plot(wisc.hclust)
abline(h= 19, col="red", lty=2)
```

## Selecting numbers of clusters

```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, k = 4)
table(wisc.hclust.clusters, diagnosis)
```

## Combining methods

The idea here is that I can take my new variables (the scores on the PCs `wisc.pr$x`) that are better descriptors of the dara-set than the original features (i. e. the 30 columns in `wisc.data`) and use these as a basis for clustering. 

```{r}
pc.dist <- dist(wisc.pr$x[ ,1:3])
pc.disthclust <- hclust(pc.dist, method = "ward.D2")
plot(pc.disthclust)
```

```{r}
grps <- cutree(pc.disthclust, k = 2)
table(grps)
```

I can now run `table()` with both my clustering `grps` and the expert `diagnosis`

```{r}
table(grps, diagnosis)
```

Our cluster "1" has 179 with "M" diagnosis 
or cluster "2" has 333 "B" diagnosis

179 TP
24 FP
333 TN
33 FN

Sensitivity: TP/(TP+FN)
```{r}
179/(179+33)
```

Specificity: TN/(TN+FP)
```{r}
333/(333+24)
```

## Prediction 

```{r}
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```

```{r}
plot(wisc.pr$x[,1:2], col=grps)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")
```

