---
title: "Class 05: Data Visualization with GGPLOT"
author: "Joli DeLucia (A17320476)"
format: pdf
toc: true
---


## Background

There are lot's of ways to make plots in R. These include so-called "base R" (like the `plot()`) and add on packages like **ggplot2**.

Let's make the same plot with these two graphics systems. We can use the inbuilt `cars` data set:

```{r}
head( cars )
```

With "base R" we can simply:

```{r}
plot(cars)
```

Now let's try ggplot. First I need to install the package using `install.packages("ggplot2")`

> **N.B.** We never run an `install.packages()` in a code chunk otherwise we will re-install needlessly every time we render our document.

Everytime you want to use an add-on package we need to load it up with a call to `library()`

```{r}
library(ggplot2)
ggplot(cars)
```

Every ggplot needs at least 3 things:

1.  The **data** i.e. stuff to plot as a data.frame
2.  The **aes** or aesthetics that map the data to the plot
3.  The **geom** or geometry i.e the plot type such as points, lines, etc.

```{r}
ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point()+
  geom_smooth(method=lm, se = FALSE, color = "pink") +
  labs(x = "Speed(MPH)", y = "Distance (ft)", title = "Stopping Distance of Old Cars") +
  theme_bw()
```

## Gene Expression Plot

Read some data on the effect of GLP-1 inhibitor (drug) on gene expression values:

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

Version 1 plot = start simple by getting some ink on the page.

```{r}
ggplot(genes) +
  aes(x = Condition1, y = Condition2, col=State) +
  geom_point() +
  scale_color_manual(values = c("purple", "pink", "red")) +
  labs(title = "Effect of GLP-1 inhibitors on Gene Expression", x = "Control (no drug)", y = "GLP-1 Inhibitor (drug)") +
  theme_light()
```

Let's color by `State` up, down or no change

```{r}
genes$State
```

## Going Further with gapminder

Here we explore the famous `gapminder` dataset with some custom plots.

```{r}
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
head(gapminder)
```

> Q. How many rows does this dataset have?

```{r}
nrow(gapminder)
```

> Q. How many different continents are in this dataset?

```{r}
table(gapminder$continent)
```

Version 1 plot GDP vs lifeExp for all rows

```{r}
ggplot(gapminder) +
  aes(gdpPercap, lifeExp, col= continent) +
  geom_point()
```

I want to see a plot for each continent - in ggplot lingo this is called "faceting"

```{r}
ggplot(gapminder) +
  aes(gdpPercap, lifeExp, col= continent,) +
  geom_point() +
  facet_wrap(~continent)
```

## First look at the dplyr package

Another add-on package with a function called `filter()` that we want to use.

```{r}
library(dplyr)
```

```{r}
filter(gapminder, year == 2007, country =="Ireland")
```

```{r}
input <- filter(gapminder, year == 2007 | year == 1977)
```

```{r}
ggplot(input) +
  aes(gdpPercap, lifeExp, col = continent) +
  geom_point() +
  facet_wrap(~year)

```

