---
title: "homework06"
author: "Joli DeLucia (A17320476)"
format: pdf
---

## Creating a Function

> Q.6 How would you generalize the original code above to work with any set of input
protein structures?

Before creating a function code:

```{r}
library(bio3d)
s1 <- read.pdb("4AKE") # kinase with drug
s2 <- read.pdb("1AKE") # kinase no drug
s3 <- read.pdb("1E4Y") # kinase with drug
s1.chainA <- trim.pdb(s1, chain="A", elety="CA")
s2.chainA <- trim.pdb(s2, chain="A", elety="CA")
s3.chainA <- trim.pdb(s1, chain="A", elety="CA")
s1.b <- s1.chainA$atom$b
s2.b <- s2.chainA$atom$b
s3.b <- s3.chainA$atom$b
plotb3(s1.b, sse=s1.chainA, typ="l", ylab="Bfactor")
plotb3(s2.b, sse=s2.chainA, typ="l", ylab="Bfactor")
plotb3(s3.b, sse=s3.chainA, typ="l", ylab="Bfactor")
```

Process of creating function code:

```{r}
library(bio3d)
analyze_protein <- function(pdb_id) {
  pdb <- read.pdb(pdb_id)
  pdb_chainA <- trim.pdb(pdb, chain="A", elety="CA")
  bvals <- pdb_chainA$atom$b
  plotb3(bvals, sse=pdb_chainA, typ="l", ylab="B-factor")
}
```

Testing it out:

```{r}
analyze_protein("4AKE")
analyze_protein("1AKE")
analyze_protein("1E4Y")
```

## Inputs

1. `read.pdb(pbd.id)` is for reading the structure from the protein accession code within the PDB.
2. `trim.pdb()` is for trimming the structure to the chain and atom type. The default chain read is "A" and the default atom type to keep is "CA"
3. `pdb_chainA$atom$b` extracts the B-factors from the trimmed structure
4. `plotb3()` makes the plot for the B-factor. The `sse=` part puts the secondary structures like in the original code before the function. 

## Outputs

1. Primary output is the B-factor line plot for the structure inputed