---
title: "Class06: R Functions"
author: "Joli DeLucia (A17320476)"
format: pdf
toc: true
---

## Background

Functions are at the heart of using R. Everything we do involves calling and using functions (from data, input, analysis to results output).

All functions in R have at least 3 things: 

1. A **name** the thing we use to call the function.
2. One or more input **arguments** that are comma separated
3. The **body**, lines of code between curly brackets { } that does the work of the function

## A first function

Let's write a silly function that adds some numbers:

```{r}
add <- function(x) {
  x + 1
}
```

Let's try it out

```{r}
add(100)
```

Will this work

```{r}
add( c(100,200, 300) )
```

Modify to be more useful and add more than just 1

```{r}
add <- function(x, y = 1) {
  x + y
}
```


Let's try it

```{r}
add(100, 10)
```

Will this work?

```{r}
add(100)
```

```{r}
log(10, base = 10)
```

> **N.B** Input arguments can be either **required** or **optional**. The later have a fall-back default that is specified in the function code with the equals sign =. 

```{r}
#add(100, 200, 300)
```

## A second function

All functions in R look like this

```
name <- function(arg) {
  body
}
```

The `sample()` function in R ... 

```{r}
sample(1:10, 4)
```

> Q. Return 12 numbers picked randomly from the input 1:10

```{r}
sample(1:10, 12, replace = TRUE)
```

> Q. Write the code that generates a random 12 nucleotide long DNA sequence?

```{r}
sample( c("A", "T", "G", "C"), 12, replace = TRUE)
```

> Q. Write a first version function called `generate_dna()` that generates a user specified length `n` random DNA sequence?

```{r}
generate_dna <- function(n = 6) {
  sample( c("A", "T", "G", "C"), size = n, replace = TRUE)
}
```

```{r}
generate_dna(15)
```

> Q. Modify your function to return a FASTA like sequence so rather than [1] "G" "A" "C" "G" we want "GACG"

```{r}
generate_dna <- function(n = 6) {
  paste(sample( c("A", "T", "G", "C"), size = n, replace = TRUE), collapse = "")
}
```

```{r}
generate_dna(15)
```

> Q. Give the user the option to return the FASTA format output or the standard multi-element vector format?

```{r}
generate_dna <- function(n = 6, fasta = TRUE) {
  ans <- sample( c("A", "T", "G", "C"), size = n, replace = TRUE)
  
  if(fasta) {
    ans <- paste(ans, collapse = "")
  }
  
  return(ans)
}
```

```{r}
generate_dna(15, fasta = F)
```
```{r}
generate_dna(15, fasta = T)
```


```{r}
generate_dna(15)
```

## A new cool function

> Q. Write a function called `generate_protein()` that generates a user specified length protein sequence in FASTA like format?

```{r}
aa20 <- c("A","R","N","D","C","Q","E","G","H","I",
          "L","K","M","F","P","S","T","W","Y","V")
generate_protein <- function(n =6) {
  paste(sample(aa20, size = n, replace = T), collapse = "")
}
```

```{r}
generate_protein()
```

> Q. Use your new `generate_protein()` function to generate all sequences between length 6 and 12 amino acids in length and check if any of these are unique in nature (i.e found in the NR database at NCBI)?

```{r}
generate_protein(6)
generate_protein(7)
generate_protein(8)
generate_protein(9)
generate_protein(10)
generate_protein(11)
generate_protein(12)
```

Or we could do a `for()` loop:

```{r}
for(i in 6:12) {
  cat(">", i, "\n", sep = "")
  cat ( generate_protein(i), "\n" )
}
```

